How to select predictive models for decision-making or causal inference?

### 2023-12-17 how_to_select_causal_models.tex
iter

\paragraph{Nuisance models: more gain than pain}
%
Predictive models are increasingly used to reason about causal effects, for
instance in precision medicine to drive individualized decision. Our results
highlight that they should be selected, validated, and tuned using different
procedures and error measures than those classically used to assess prediction
(estimating the so-called $\mu\text{-risk}$). Rather, selecting the best outcome
model according to the $R\text{-risk}$ (eq.\, Definition \ref{def:r_risk}) leads
to more valid causal estimates. Estimating the $R\text{-risk}$ requires a more
complex procedure than standard cross-validation used \eg in machine
learning: it involves fitting nuisance models necessary for model evaluation,
though our results show that these can be learned on the same set of data as the
outcome model evaluated. The nuisance models must be well estimated (Figure
\ref{fig:all_datasets_nuisances_comparison}). Our results show that using for
nuisance models a flexible stacking-based family of estimator suffices for good
model selection. For propensity score, we used the Brier score,
which is minimized by the true individual probability. An easy mistake is to
use calibration errors popular in machine learning
\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
as these select not for the individual posterior probability but for an
aggregate error rate \cite{perez2022beyond}.

@@

\paragraph{Extension to binary outcomes}
While we focused on continuous outcomes, in medicine, the target outcome
is often a categorical variable such as mortality status or diagnosis. In
this case, it may be interesting to focus on other estimands than the
Average Treatment Effect $\mathbb{E}[Y(1)] -\mathbb{E}[Y(0)] $, for instance the relative risk
$\frac{\mathbb P(Y(1) = 1)}{\mathbb P(Y(0) = 1)}$
or the odd ratio, $\frac{\mathbb P(Y(1) = 1) / [1 - \mathbb P(Y(1)
    =1)]}{\mathbb P(Y(0) = 1) / [1 - \mathbb P(Y(0) = 1]}$ are often used
\cite{austin2017estimating}. While the odds ratio is natural for
case-control studies \cite{rothman2008case}, other measures
can reduce heterogeneity \cite{colnet2023risk}.
In the log domain, the ratios are written as a difference, the
framework studied here can directly apply.
% In particular, the log odds ratio is estimated by the common
% cross-entropy loss (or log loss) as in logistic regression.

>>>>

\paragraph{Nuisance models: more gain than pain}
%
Predictive models are increasingly used to reason about treatment effects, for
instance in precision medicine to drive individualized decision. Our results
highlight that they should be selected, validated, and tuned using different
procedures and error measures than those classically used to assess prediction. Rather, selecting the best outcome
model according to the $R\text{-risk}$ (eq.\, Definition \ref{def:r_risk}) leads
to more valid causal estimates. Estimating the $R\text{-risk}$ requires a more
complex procedure than standard cross-validation used \eg in machine
learning: it involves fitting nuisance models necessary for model
evaluation.
Our results show that these can be learned on the same set of data as the
outcome models evaluated. The nuisance models must be well estimated (Figure
\ref{fig:all_datasets_nuisances_comparison}). Our results show that using for
nuisance models a flexible stacking-based family of estimator suffices for good
model selection. To select propensity score models, we used the Brier score,
minimized by the true individual probability. An easy mistake is to
use calibration errors popular in machine learning
\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
as these select not for the individual posterior probability but for an
aggregate error rate \cite{perez2022beyond}.

@@

%\paragraph{Extension to binary outcomes}
%While we focused on continuous outcomes, in medicine, the target outcome
%is often a categorical variable such as mortality status or diagnosis. In
%this case, it may be interesting to focus on other estimands than the
%Average Treatment Effect $\mathbb{E}[Y(1)] -\mathbb{E}[Y(0)] $, for instance the relative risk
%$\frac{\mathbb P(Y(1) = 1)}{\mathbb P(Y(0) = 1)}$
%or the odd ratio, $\frac{\mathbb P(Y(1) = 1) / [1 - \mathbb P(Y(1)
%    =1)]}{\mathbb P(Y(0) = 1) / [1 - \mathbb P(Y(0) = 1]}$ are often used
%\cite{austin2017estimating}. While the odds ratio is natural for
%case-control studies \cite{rothman2008case}, other measures
%can reduce heterogeneity \cite{colnet2023risk}.
%In the log domain, the ratios are written as a difference, the
%framework studied here can directly apply.
%% In particular, the log odds ratio is estimated by the common
%% cross-entropy loss (or log loss) as in logistic regression.
